Reconstructing Propositional Proofs in Type Theory

### 2017-10-26 paper/sections/introduction.tex
[ paper ] fixed some corrections.

Reconstructing and verifying proofs delivered by automatic theorem provers
(henceforth ATP) are non-trivial tasks.
In the last decades, \ATPs is fast becoming a key instrument in 
different disciplines (\eg~verifying cryptographic protocols for information security) and researchers have shown an increased interest to prove the validity of their deduction algorithms.
One reason to say that is the bugs found in many ATPs~\cite{Keller2013,Bohme2011,Fleury2014}.
For that reason, in order to give confidence to the user of these systems,
many ATPs has started to include in their outputs, the proof of their
conclusions. However, existing research recognizes that in many cases these proofs encode non-trivial reasoning
hard to reconstruct and therefore hard to verify. The proof-reconstruction
address this problem and becomes in some cases a reverse engineering of the
prover reasoning.

Therefore, the proof replied by the prover and its quality play an important role for proof-reconstruction. To verify such a proof, it is necessary to have
it in a consistent format, that is, a full script 
describing step-by-step with exhaustive details and without ambiguities, the derivation to get the proof.
For classical propositional logic (henceforth \CPL) from a list of at least forty\footnote{\ATPs 
available from the web
service \name{SystemOnTPTP} of the TPTP World.} \ATPs, just a few
provers were able to deliver proofs and fewer like \name{EProver} or \Metis 
replied the proof in an admissible file format.

One approach to address the proof-reconstruction 
is integrating the \ATP, the \emph{source} system, with a proof assistant, the 
\emph{target} system.
The target system becomes in the proof checker. These proof assistants allow 
us to define the formal language used in the proofs, \ie, operators, logic 
variables, axioms, and theorems. A proof-reconstruction tool provides such an
integration, translating the replied generated by the prover into the formalism of the proof assistant.

Previous studies have reported proof-reconstruction using proof assistants for classical logic where the development is at a mature stage.
Even though we choose a proof assistant for type theory
where proof verification becomes a type-checking task of the proof-term following the Curry-Howard isomorphism. Therefore the proof-reconstruction tool generates proof-terms by reconstructing the proof using the formalization 
the prover in type theory. However, this approach suffers from some limitations of type theory as we describe in Section~\ref{sec:limitations-type-theory}.

In this paper, we describe how to reconstruct proofs in type theory
replied by \Metis prover\footnote{v2.3 release 2017102}.
We ported the subset of the \Metis inference rules for the propositional 
fragment using a syntactical treatment of the logic. The \Metis reasoning was formalized in \Agda in~\cite{AgdaProp,AgdaMetis}. We built a proof-reconstruction tool named \Athena~\cite{Athena} that is able to generate \Agda\footnote{v2.5.3} proof-terms for \Metis derivations. We have avoided the use of propositions meanings towards a future work to support other logics
but also contrast the usage of a deep embedding of the logic with
a shallow embedding.

This paper has been organized in the following way.
In Section~\ref{sec:limitations-type-theory} some limitations of type theory
are discussed for proof-reconstruction point view.
In Section~\ref{sec:metis-language-and-proofs}, we introduce the
\Metis prover and its deduction system.
In Section~\ref{sec:proof-reconstruction}, we show our
approach to reconstruct proofs deliver by \Metis.
Related work is described in Section~\ref{sec:related-work}.
Main findings and suggestion for future work is presented Section~\ref{sec:conclusions}.

@@

The program and formalizations in \Agda mentioned in this paper are available
as \verb!Git! repositories at \verb!Github!:

\begin{itemize}
  \item The \Athena program to translate proofs generated by \Metis to \Agda code:
  \url{http://github.com/jonaprieto/athena}.
  \item The \verb!agda-prop! library as a formal system in \Agda for propositional logic:
  \url{http://github.com/jonaprieto/agda-prop}.
  \item The \verb!agda-metis! library as a formalization in \Agda to justify \Metis proofs in propositional
  logic:
  \url{http://github.com/jonaprieto/agda-prop}.
\end{itemize}

>>>>

An automatic theorem provers (henceforth \ATP) is a tool that intends to prove
conjectures based on assumptions and hypothesis in a logic specification.
Verifying proofs generated by \ATPs are non-trivial tasks.
In the last decades, \ATPs are fast becoming a key instrument in
different disciplines and real applications (\eg, verifying railway interlocking systems in~\cite{Kanso2012}).
Therefore, researchers from academy and industry have shown an increased interest to prove the validity of their deduction algorithms.

See for instance bugs found in ATPs in~\cite{Keller2013,Bohme2011,Fleury2014}.
For that reason, in order to give confidence to the user of these systems,
many ATPs have started to include in their outputs, a witness of their
conclusions. However, existing research recognizes that in many cases these proofs encode non-trivial reasoning
hard to reconstruct and therefore hard to verify. The proof-reconstruction
address this problem.
It becomes in mostly cases a reverse engineering to justify the prover reasoning.
That happen when the reconstruction is made by another and not by
the developers of the \ATP.

Therefore, the proof generated by the prover and its presentation play an important role for proof-reconstruction. To verify such a proof, it is necessary to have
it in a consistent format, that is, a full script
describing step-by-step with exhaustive details and without ambiguities, the derivation to get the proof.
For classical propositional logic (henceforth \CPL) from a list of at least forty\footnote{\ATPs
available from the web
service \name{SystemOnTPTP} of the TPTP World.} \ATPs, just a few
provers were able to output proofs.

One approach to address the proof-reconstruction
is integrating the \ATP, the \emph{source} system, with a proof-assistant, the
\emph{target} system.
The target system becomes in the proof checker. These proof-assistants allow
us to define the formal language used in the proofs, \ie, operators, logic
variables, axioms, and theorems. A proof-reconstruction tool provides such an
integration, translating the replied generated by the prover into the formalism of the proof-assistant.

Previous studies have reported proof-reconstruction using proof-assistants
based on classical logic where the development is at a
mature stage~\cite{paulson2010three,hurlin07practical,kaliszyk2013}.

Another approaches has been proposed for proof-reconstructing using
the formalism of type theory.
The reader can review~\cite{Bezem2002,armand2011,Ekici2017,kanso2016light}.

In this paper, we describe how to reconstruct proofs in type theory
generated by the \Metis prover\footnote{v2.3 release 2017102}~\cite{hurd2003first}.
We ported the subset of the \Metis inference rules for the propositional
fragment using a syntactical treatment of the logic. The \Metis reasoning was
formalized in \Agda~\cite{agdateam} in~\cite{AgdaProp,AgdaMetis}. We built a
proof-reconstruction tool named \Athena~\cite{Athena} written in \Haskell
that is able to generate \Agda proof-terms for \Metis derivations.

This paper has been organized in the following way.
In Section~\ref{sec:limitations-type-theory} some limitations of type theory
are discussed from proof-reconstruction point view.
In Section~\ref{sec:metis-language-and-proofs}, we introduce the
\Metis prover.
In Section~\ref{sec:proof-reconstruction} we show our
approach to reconstruct \Metis proofs.
Related work is described in Section~\ref{sec:related-work}.
Conclusions and suggestions for future work are  presented in
Section~\ref{sec:conclusions}.

@@

The program and formalizations in \Agda and \Haskell mentioned in this paper are available
as \verb!Git! repositories at \verb!Github!:

\begin{itemize}
  \item The \Athena program that translates proofs generated by \Metis to \Agda:
  \url{http://github.com/jonaprieto/athena}.
  \item The \verb!agda-prop! library as a formalization in \Agda for propositional logic:
  \url{http://github.com/jonaprieto/agda-prop}.
  \item The \verb!agda-metis! library as a formalization in \Agda to justify \Metis proofs of propositional
  logic:
  \url{http://github.com/jonaprieto/agda-prop}.
\end{itemize}

### 2017-10-26 paper/sections/limitations-type-theory.tex
[ paper ] fixed some corrections.

% In Section~\label{ssec:structural-recursion},
% we describe how this observation is an issue from the type theory point of
% view. In Section~\ref{ssec:intuitionistic-logic} we mention
% the use intuitionistic-logic and its limitation.

@@

Since we pretend to reconstruct proofs replied by \Metis, and \Metis works
in classical logic,  we have included as an axiom, the
principle of excluded third (\abbre{PEM}).

>>>>

Even though we choose a proof-assistant for type theory
where proof verification becomes a type-checking task of the proof-term following the Curry-Howard isomorphism.

Therefore the proof-reconstruction tool generates proof-terms by reconstructing the proof using the formalization
the prover in type theory. However, this approach suffers from some limitations of type theory as we describe in Section~\ref{sec:limitations-type-theory}.

@@

Since we pretend to reconstruct proofs generated by \Metis, and \Metis works
in classical logic,  we have included as an axiom, the
principle of excluded third (\abbre{PEM}).

### 2017-10-26 paper/sections/metis-language-and-proofs.tex
[ paper ] fixed some corrections.

A proof delivered by \Metis encodes a natural
deduction proof.
With these inference rules as the only valid deduction steps, \Metis
attempts to prove conjectures by refutation (\eg,
\emph{falsium} in the root of the \TSTP derivation).

These proofs are directed acyclic graphs, trees of refutations
Each node stands for an application of an inference rule and the leaves
in the tree represent formulas in the given problem. Each node is
labeled with a name of the inference rule (\eg, \canonicalize).
Each edge links a premise with one conclusion.
The proof graphs have at their root the conclusion
$⊥$, since \Metis delivers proofs by refutation.

>>>>

A proof generated by \Metis encodes a natural
deduction proof.
With these inference rules as the only valid deduction steps, \Metis
attempts to prove conjectures by refutation (\eg,
\emph{falsium} in the root of the \TSTP derivation).

These proofs are directed acyclic graphs, trees of refutations
Each node stands for an application of an inference rule and the leaves
in the tree represent formulas in the given problem. Each node is
labeled with a name of the inference rule (\eg, \canonicalize).
Each edge links a premise with one conclusion.
The proof graphs have at their root the conclusion
$⊥$, since \Metis proofs are refutations.

### 2017-10-26 paper/sections/proof-reconstruction.tex
[ paper ] fixed some corrections.

Finally, we type-check the \Agda proof-term. If the type-checking
success, the \TSTP derivation delivered by \Metis is correct
module \Agda and the proposed formalizations for the
propositional logic and the \Metis inferences.
In that case \Agda outputs an interface file.
Otherwise, when type-checking fails, the failure must be
investigated by the user looking at the error in the \TSTP derivation
by \Metis, in the translation by \Athena, in the \Agda formalizations
mentioned above or in the type-checker~\Agda.

>>>>

Finally, we type-check the \Agda proof-term. If the type-checking
success, the \TSTP derivation generated by \Metis is correct
module \Agda and the proposed formalizations for the
propositional logic and the \Metis inferences.
In that case \Agda outputs an interface file.
Otherwise, when type-checking fails, the failure must be
investigated by the user looking at the error in the \TSTP derivation
by \Metis, in the translation by \Athena, in the \Agda formalizations
mentioned above or in the type-checker~\Agda.

### 2017-10-26 paper/sections/related-work.tex
[ bib ] Updated HOL reference.
[ paper ] fixed some corrections.

In type theory, for \Agda and  for the same propositional fragmented used in this paper,
we found the proof-reconstruction proposed by \citeauthor{Kanso2012}
in~\cite{Kanso2012,kanso2016light}. The author present an approach similar to
the reasoning presented in Section~\ref{ssec:workflow}.
Nonetheless, his approach includes semantics reasoning that as we mentioned
in the introduction differs completely from our proof-reconstruction approach.

@@

\citeauthor{Hurd1999}~\cite{Hurd1999} integrates the first-order resolution
prover \name{Gandalf} with the high-order theorem prover
\name{HOL}~\cite{norrish2007hol}.
Its \verb!GANDALF_TAC! tactic is able to reconstruct \name{Gandalf} proofs
by using a \abbre{LCF} model. For \name{HOL Light}, a version of
\name{HOL} but with a simpler logic core, the \SMT solver \name{CVC4}
was integrated. \citeauthor{kaliszyk2013}~\cite{kaliszyk2013}
reconstruct proofs from different \ATPs with the \name{PRocH} tool by
replaying detailed inference steps from the \ATPs with internal
inference methods implemented in \name{HOL Light}.

>>>>

In type theory, for \Agda and  for the same propositional fragmented used in this paper,
we found the proof-reconstruction proposed by \citeauthor{Kanso2012}
in~\cite{Kanso2012,kanso2016light}. The author present an approach similar to
the reasoning presented in Section~\ref{ssec:workflow}.
Nonetheless, his approach includes semantics reasoning. We have avoided the
use of propositions meanings towards a future work to support other logics but
also contrast the usage of a deep embedding of the logic with a shallow
embedding.

@@

\citeauthor{Hurd1999}~\cite{Hurd1999} integrates the first-order resolution
prover \name{Gandalf} with the high-order theorem prover
\name{HOL}~\cite{norrish2017hol}.
Its \verb!GANDALF_TAC! tactic is able to reconstruct \name{Gandalf} proofs
by using a \abbre{LCF} model. For \name{HOL Light}, a version of
\name{HOL} but with a simpler logic core, the \SMT solver \name{CVC4}
was integrated. \citeauthor{kaliszyk2013}~\cite{kaliszyk2013}
reconstruct proofs from different \ATPs with the \name{PRocH} tool by
replaying detailed inference steps from the \ATPs with internal
inference methods implemented in \name{HOL Light}.
